---
title: "Untitled"
author: "AAA"
date: "2026-06-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


# 1. Load necessary libraries
if (!require("gamlss")) install.packages("gamlss")
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("lubridate")) install.packages("lubridate")

library(gamlss)
library(tidyverse)
library(lubridate)

```

## T-student or normal
In a Student-t distribution, the parameter ν (degrees of freedom) controls the tails.
If ν is small (e.g., 3 to 10), the data has heavy tails (outliers).
If ν is large (e.g., > 30), the Student-t becomes virtually identical to a Normal distribution.
```{r check-normal}

# 2. Load and Preprocess Data
df <- read.csv("data/filtered_marathons_results.csv")

# Convert "Official.Time" (H:M:S) to total seconds and take log
df <- df %>%
  mutate(
    seconds = as.numeric(hms(Official.Time)),
    y = log(seconds),
    Sex = as.factor(M.F),
    Age = as.numeric(Age)
  ) %>%
  filter(!is.na(y)) # Remove rows with missing times

# 3. Fit the models
# Model 1: Normal Distribution
model_norm <- gamlss(y ~ Sex + pb(Age), 
                     family = NO, 
                     data = df, 
                     control = gamlss.control(n.cyc = 100))

# Model 2: Student-t Distribution (TF)
# This estimates the nu (df) parameter automatically
model_t <- gamlss(y ~ Sex + pb(Age), 
                  family = TF, 
                  data = df, 
                  control = gamlss.control(n.cyc = 100))

# 4. Compare models using AIC
# Lower AIC is better.
cat("AIC Normal:", AIC(model_norm), "\n")
cat("AIC Student-t:", AIC(model_t), "\n")

# 5. Extract the estimated nu (degrees of freedom) for the t-distribution
# If nu is small (< 20), the Student-t is significantly better.
nu_val <- exp(model_t$nu.coefficients)
cat("Estimated Degrees of Freedom (nu):", nu_val, "\n")

# 6. Diagnostic Plots
# Check if the residuals of the normal model have "heavy tails"
plot(model_norm, summary = TRUE) # Look at the QQ-plot
```

The ν (Nu) Parameter
When you fit the Student-t model, check the estimated degrees of freedom (ν):
ν<10: Your data has very heavy tails. Use Student-t.
10<ν<30: Moderate outliers. Student-t is safer.
ν>50: The distribution is effectively Normal. Use Normal.


The Q-Q Plot (The Visual Test)
Look at the Q-Q Plot generated by plot(model_norm):
If the points in the corners (the "tails") curve away from the straight diagonal line, your residuals are not normal.
If the points form an "S" shape or flare out at the ends, the Student-t will handle those outliers much better.

Why use Log(Time) with Student-t?
Marathon data (like your CSV) often contains "stragglers"—runners who take much longer than the average. Even after taking the log(), these individuals often appear as outliers in the left or right tail. The Student-t distribution is "robust," meaning these slow runners won't bias your estimate of the average pace 
μ as much as a Normal distribution would.