---
title: "Predict"
output: html_document
---

#  Prediction using Model 3

```{r}
library(readr)
library(dplyr)
library(lubridate)
library(cmdstanr)
library(posterior)
library(ggplot2)
```

# Load data

```{r}
df_raw <- read_csv("repeated_ids_with_countries.csv")
```

# Prepare data

```{r}
df_model <- df_raw %>%
  mutate(
    log_time = log(as.numeric(hms(`Official Time`)) / 3600),
    Sex = ifelse(`M/F` == "M", 1, 0),
    Age = as.numeric(scale(Age)),
    Age2 = Age^2,
    year_id = as.integer(as.factor(Year)),
    country_id = as.integer(as.factor(Country)),
    runner_id = as.integer(as.factor(ID))
  ) %>%
  filter(!is.na(log_time)) %>%
  mutate(Year = as.integer(Year))
```

# Train-test split

```{r}
df_train <- df_model %>%
  filter(Year <= 2018)

df_test <- df_model %>%
  filter(Year == 2019)
```

# Keep runners observed in both train and test

```{r}
common_runners <- intersect(
  unique(df_train$runner_id),
  unique(df_test$runner_id) )

df_pred <- df_test %>%
  filter(runner_id %in% common_runners)
```

# Load fitted Model 3

```{r}
fit3 <- readRDS("fit_model3.rds")
```

# Extract posterior means

```{r}
main_params <- fit3$summary(c(
  "beta0",
  "beta_sex",
  "beta_age",
  "beta_age2"
))

b0 <- main_params$mean[main_params$variable == "beta0"]
bsex <- main_params$mean[main_params$variable == "beta_sex"]
bage <- main_params$mean[main_params$variable == "beta_age"]
bage2 <- main_params$mean[main_params$variable == "beta_age2"]
```

# Extract runner effects

```{r}
runner_effects <- fit3$summary("alpha_runner") %>%
  mutate(
    runner_id = as.integer(
      gsub("alpha_runner\\[|\\]", "", variable)
    )
  ) %>%
  select(runner_id, alpha_runner = mean)
```

#  Extract country effects

```{r}
country_effects <- fit3$summary("alpha_country") %>%
  mutate(
    country_id = as.integer(
      gsub("alpha_country\\[|\\]", "", variable)
    )
  ) %>%
  select(country_id, alpha_country = mean)
```

# Extract year effect for 2019

```{r}
year_effects <- fit3$summary("alpha_year") %>%
  mutate(
    year_id = as.integer(
      gsub("alpha_year\\[|\\]", "", variable)
    )
  ) %>%
  select(year_id, alpha_year = mean)
```

# Build predictions

```{r}
df_pred <- df_pred %>%
  left_join(runner_effects, by = "runner_id") %>%
  left_join(country_effects, by = "country_id") %>%
  left_join(year_effects, by = "year_id") %>%
  mutate(
    pred_log_time =
      b0 +
      bsex * Sex +
      bage * Age +
      bage2 * Age2 +
      alpha_runner +
      alpha_country +
      alpha_year,
    
    observed_hours = exp(log_time),
    predicted_hours = exp(pred_log_time),
    
    observed_minutes = observed_hours * 60,
    predicted_minutes = predicted_hours * 60,
    
    error_minutes = predicted_minutes - observed_minutes,
    abs_error_minutes = abs(error_minutes)
  )
```

# Prediction metrics

```{r}
prediction_metrics <- df_pred %>%
  summarise(
    n_predictions = n(),
    MAE_minutes = mean(abs_error_minutes, na.rm = TRUE),
    RMSE_minutes = sqrt(mean(error_minutes^2, na.rm = TRUE)),
    correlation = cor(observed_minutes, predicted_minutes)
  )

print(prediction_metrics)
```

# Performance level error

```{r}
# Does it predict fast and slow runners equally well?
df_pred %>%
  mutate(
    group = case_when(
      observed_minutes < 180 ~ "Sub 3h",
      observed_minutes < 240 ~ "3h-4h",
      TRUE ~ "Over 4h"
    )
  ) %>%
  group_by(group) %>%
  summarise(
    MAE = mean(abs_error_minutes),
    RMSE = sqrt(mean(error_minutes^2)),
    n = n()
  )
```

# Top 20 best predictions

```{r}
df_pred %>%
  arrange(abs_error_minutes) %>%
  select(
    Name,
    observed_minutes,
    predicted_minutes,
    abs_error_minutes
  ) %>%
  head(20)
```

#  Worst predictions

```{r}
df_pred %>%
  arrange(desc(abs_error_minutes)) %>%
  head(20)
```

# Predictive R²

```{r}
correlation = 0.92
correlation^2
```

# Error vs. real-time graph

```{r}
ggplot(
  df_pred,
  aes(
    x = observed_minutes,
    y = error_minutes
  )
) +
  geom_point(alpha = 0.3) +
  geom_hline(
    yintercept = 0,
    colour = "red"
  ) +
  theme_minimal()
```

# Save predictions

```{r}
write_csv(df_pred, "model3_predictions_2019.csv")
write_csv(prediction_metrics, "model3_prediction_metrics.csv")
```

# Plot observed vs predicted

```{r}
ggplot(df_pred, aes(x = observed_minutes, y = predicted_minutes)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(
    title = "Observed vs Predicted Marathon Finish Times",
    x = "Observed time (minutes)",
    y = "Predicted time (minutes)"
  ) +
  theme_minimal()

ggsave(
  "model3_observed_vs_predicted.png",
  width = 7,
  height = 5,
  dpi = 300
)
```

# Plot prediction errors

```{r}
ggplot(df_pred, aes(x = error_minutes)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  labs(
    title = "Prediction Error Distribution",
    x = "Prediction error (minutes)",
    y = "Count"
  ) +
  theme_minimal()

ggsave(
  "model3_prediction_errors.png",
  width = 7,
  height = 5,
  dpi = 300
)
```
